---
title: "Heat_wave_temperature"
author: "Wendel Raymond"
date: "July 28, 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Analytical code for Raymond et al. Ecology
List of Authors: Wendel W. Raymond, Julie S. Barber, Megan N. Dethier, Hilary A. Hayford, Christopher D.G. Harley, Teri L. King, Blair Paul, Camille A. Speck, Elizabeth D. Tobin, Ann E.T. Raymond, P. Sean McDonald

Title: Assessment of the impacts of an unprecedented heatwave on intertidal shellfish of the Salish Sea

Journal: Ecology 

## Heat wave temperature data
The code presented here queries and processes temperature and water level data from online databases and static .csv files included in this repository. 

```{r libraries}
library(noaaoceans)
library(tidyverse)
library(cowplot)
library(lubridate)
library(rnoaa)
library(daymetr)
library(heatwaveR)
library(tiff)

theme_set(theme_classic())
```


### Load Data
Import NOAA and weather station data as applicable.

```{r load data}
## Air Temp ##

# NOAA Air temp data from Toke Point, Port Townsend #
noaa.air.stations <- c("9440910", "9444900")
air.temp <- data.frame()
for (i in noaa.air.stations) {
    query.df2 <- query_coops_data(station_id = i,
                                 start_date = '20210624',
                                 end_date = '20210701',
                                 data_product = 'air_temperature',
                                 units = "metric", 
                                 time_zone = "lst_ldt",
                                 interval = 'h') 
    air.temp <- air.temp %>% bind_rows(., query.df2)
}

# Air temp from Evergreen State College #
oly.air <- read.csv("data/ESC__Weather_Station_Hourly_June2021.csv", header = TRUE, stringsAsFactors = FALSE)
oly.air$TIMESTAMP <- as.POSIXct(oly.air$TIMESTAMP, format = "%m/%d/%Y %H:%M")
oly.air$station.name <- "Budd Inlet"

# Air temp from Vancouver Intnl #
van.air <- read.csv("data/VancouverIntnl_June_AirTemp.csv", header = TRUE, stringsAsFactors = FALSE)
van.air$Date.Time..LST. <- as.POSIXct(van.air$Date.Time..LST., format = "%Y-%m-%d %H:%M")
van.air$station.name <- "Vancouver"

# Air temp from Tatoosh Island #
tat.air <- read.csv("data/Tatoosh_Weather_Heatwave_2021_wx.csv", header = TRUE, stringsAsFactors = FALSE)
tat.air$Date_Time <- as.POSIXct(tat.air$Date_Time, format = "%m/%d/%Y %H:%M")
tat.air$station.name <- "Neah Bay"

cal.air <- read.csv("data/Calvert_PruthBay_wx.csv", header = TRUE, stringsAsFactors = FALSE)
cal.air$date_time <- as.POSIXct(cal.air$date_time, format = "%m/%d/%Y %H:%M")
cal.air$station.name <- "Pruth Bay"

## Water Levels ##

# NOAA Water levels from Toke Point, Port Townsend, Neah Bay, and Budd Inlet #
noaa.tide.stations <- c("9440910", "9444900", "9443090", "9446807")
water.level <- data.frame()
for (i in noaa.tide.stations) {
    query.df3 <- query_coops_data(station_id = i,
                                 start_date = '20210624',
                                 end_date = '20210701',
                                 data_product = 'predictions',
                                 units = "metric",
                                 datum = "MLLW",
                                 interval = "h",
                                 time_zone = "lst_ldt") 
    water.level <- water.level %>% bind_rows(., query.df3)
}

# Calvert Island (Pruth Bay) tide station data #
cal.tide <- read.csv("data/PruthBay_Tide_June_heatwave.csv", header = TRUE, stringsAsFactors = FALSE)
cal.tide$TIME_TAG.PST..Z.8. <- as.POSIXct(cal.tide$TIME_TAG.PST..Z.8., format = "%Y-%m-%d %H:%M")

cal.tide <- cal.tide %>% 
  filter(!is.na(PREDICTION))

colnames(cal.tide) <- c("station.name", "t", "v", "vega")
cal.tide$date <- as.Date(cal.tide$t, "%Y-%m-%d %H:%M", tz = "America/Los_Angeles")

cal.tide <- cal.tide[, c(1, 2, 3, 1)]
colnames(cal.tide) <- c("station", "t", "v", "station.name")

# Canada Vancouver tide station data #
van.tide <- read.csv("data/Vancouver_Tide_June_heatwave_2.csv", header = TRUE, stringsAsFactors = FALSE)
van.tide$TIME_TAG.PST..Z.8. <- as.POSIXct(van.tide$TIME_TAG.PST..Z.8., format = "%Y-%m-%d %H:%M")

van.tide <- van.tide %>% 
  filter(!is.na(PREDICTION))

colnames(van.tide) <- c("station.name", "t", "v", "vega")
van.tide$date <- as.Date(van.tide$t, "%Y-%m-%d %H:%M", tz = "America/Los_Angeles")

van.tide <- van.tide[, c(1, 2, 3, 1)]
colnames(van.tide) <- c("station", "t", "v", "station.name")
```

### Load Gridded climate data
Pull from Daymet 2 X 2 degree surface grid. https://daymet.ornl.gov/getdata

```{r gridded}
## Olympia ##
oly.grid <- download_daymet(site = "olympia", lat = 47.064316, lon = -122.910577, start = 1991, end = 2020, internal = TRUE)

## Port Townsend + Vancouver BC ##
ptv.grid <- download_daymet(site = "port_townsend", lat = 48.1172, lon = -122.789, start = 1991, end = 2020, internal = TRUE)

## Neah Bay ##
nhb.grid <- download_daymet(site = "neah_bay", lat = 48.377, lon = -124.6897, start = 1991, end = 2020, internal = TRUE)

## Toke Point ##
tkp.grid <- download_daymet(site = "toke_point", lat = 46.7499, lon = -124.025, start = 1991, end = 2020, internal = TRUE)

## Calvert Island ##
cvi.grid <- download_daymet(site = "calvert_island", lat = 51.6271, lon = -127.645, start = 1991, end = 2020, internal = TRUE)

```

### Data managment
#### Stations
clean up for naming and format consistency
```{r mgmt stations}
station_names <- data.frame(station = c("9440910", "9444900", "9443090", "9446807"),
                            station.name = c("Toke Point", "Port Townsend", "Neah Bay", "Budd Inlet"))
## Air Temperature ##
air.temp <- air.temp %>% 
  mutate(v = as.numeric(v), t = as.POSIXct(t, format  = "%Y-%m-%d %H:%M" ))
air.temp <- merge(air.temp, station_names)
air.temp <- air.temp[, c(2, 3, 5)]

oly.air <- oly.air[, c(1, 3, 23)]
colnames(oly.air) <- colnames(air.temp)
oly.air$t <- as.POSIXct(oly.air$t, format  = "%Y-%m-%d %H:%M:%S")

van.air <- van.air[, c(5, 10, 31)]
colnames(van.air) <- colnames(air.temp)
van.air$t <- as.POSIXct(van.air$t, format  = "%Y-%m-%d %H:%M:%S")

tat.air <- tat.air[, c(2, 3, 9)]
colnames(tat.air) <- colnames(air.temp)
tat.air$t <- as.POSIXct(tat.air$t, format = "%Y-%m-%d %H:%M:%S")

cal.air <- cal.air[, c(1, 8, 21)]
colnames(cal.air) <- colnames(air.temp)
cal.air$t <- as.POSIXct(cal.air$t, format = "%Y-%m-%d %H:%M:%S")

air.temp <- rbind(air.temp, oly.air)
air.temp <- rbind(air.temp, van.air)
air.temp <- rbind(air.temp, tat.air)
air.temp <- rbind(air.temp, cal.air)

air.temp.redu <- air.temp %>% 
  filter(t >= "2021-06-24" & t < "2021-07-01")

air.temp.redu$station.name.f <- factor(air.temp.redu$station.name, levels = c("Pruth Bay", "Vancouver", "Neah Bay", "Port Townsend", "Toke Point", "Budd Inlet"))

## Water Levels ##
water.level <- water.level %>% 
  mutate(v = as.numeric(v), t = as.POSIXct(t))

# Make usable dataframe #
water.level <- merge(water.level, station_names)
water.level <- water.level[, c(1, 3, 4, 2)]
water.level <- rbind(water.level, van.tide[])
water.level <- rbind(water.level, cal.tide)

water.level$t <- as.POSIXct(water.level$t, "%Y-%m-%d %H:%M", tz = "America/Los_Angeles")
water.level$date <- as.Date(water.level$t, "%Y-%m-%d %H:%M", tz = "America/Los_Angeles")

water.level.redu <- water.level %>% 
  filter(date >= "2021-06-24" & date < "2021-07-01") %>% 
  mutate(v = as.numeric(v))

water.level.redu$station.name <- ifelse(water.level.redu$station.name == "PRUTH BAY", "Pruth Bay", water.level.redu$station.name) 
water.level.redu$station.name <- ifelse(water.level.redu$station.name == "VANCOUVER HARBOUR", "Vancouver", water.level.redu$station.name) 

water.level.redu$station.name.f <- factor(water.level.redu$station.name, levels = c("Pruth Bay", "Vancouver", "Neah Bay", "Port Townsend", "Toke Point", "Budd Inlet"))
```

#### Exposure times
Extract times at each location where water was <= 1m. Will need to convert Canadian data (Vancouver and Pruth Bay) to MLLW.

```{r mgmt exposure}
water.level.exp <- water.level.redu
water.level.exp$v <- ifelse(water.level.exp$station.name == "Pruth Bay", water.level.exp$v - 1.01, water.level.exp$v)
water.level.exp$v <- ifelse(water.level.exp$station.name == "Vancouver", water.level.exp$v - 1.22, water.level.exp$v)

water.level.exp <- water.level.exp %>% 
  filter(v <= 1) %>% 
  filter(date == "2021-06-26" | date == "2021-06-27" | date == "2021-06-28") %>% 
  group_by(station.name.f, date) %>% 
  summarise(start = min(t),
            end = max(t))

water.level.exp[1, 3] <- as.POSIXct("2021-06-26 05:10:00")
water.level.exp[1, 4] <- as.POSIXct("2021-06-26 11:10:00")
water.level.exp[2, 3] <- as.POSIXct("2021-06-27 05:55:00")
water.level.exp[2, 4] <- as.POSIXct("2021-06-27 11:55:00")
water.level.exp[3, 3] <- as.POSIXct("2021-06-28 06:45:00")
water.level.exp[3, 4] <- as.POSIXct("2021-06-28 12:35:00")

water.level.exp[7, 3] <- as.POSIXct("2021-06-26 05:30:00")
water.level.exp[7, 4] <- as.POSIXct("2021-06-26 12:50:00")
water.level.exp[8, 3] <- as.POSIXct("2021-06-27 06:18:00")
water.level.exp[8, 4] <- as.POSIXct("2021-06-27 13:35:00")
water.level.exp[9, 3] <- as.POSIXct("2021-06-28 07:05:00")
water.level.exp[9, 4] <- as.POSIXct("2021-06-28 14:15:00")

water.level.exp[10, 3] <- as.POSIXct("2021-06-26 08:32:00")
water.level.exp[10, 4] <- as.POSIXct("2021-06-26 15:57:00")
water.level.exp[11, 3] <- as.POSIXct("2021-06-27 09:23:00")
water.level.exp[11, 4] <- as.POSIXct("2021-06-27 16:37:00")
water.level.exp[12, 3] <- as.POSIXct("2021-06-28 10:22:00")
water.level.exp[12, 4] <- as.POSIXct("2021-06-28 17:10:00")

water.level.exp[13, 3] <- as.POSIXct("2021-06-26 05:28:00")
water.level.exp[13, 4] <- as.POSIXct("2021-06-26 12:45:00")
water.level.exp[14, 3] <- as.POSIXct("2021-06-27 06:46:00")
water.level.exp[14, 4] <- as.POSIXct("2021-06-27 13:30:00")
water.level.exp[15, 3] <- as.POSIXct("2021-06-28 07:35:00")
water.level.exp[15, 4] <- as.POSIXct("2021-06-28 14:10:00")

water.level.exp[16, 3] <- as.POSIXct("2021-06-26 10:22:00")
water.level.exp[16, 4] <- as.POSIXct("2021-06-26 16:40:00")
water.level.exp[17, 3] <- as.POSIXct("2021-06-27 11:18:00")
water.level.exp[17, 4] <- as.POSIXct("2021-06-27 17:20:00")
water.level.exp[18, 3] <- as.POSIXct("2021-06-28 12:16:00")
water.level.exp[18, 4] <- as.POSIXct("2021-06-28 17:54:00")
```

#### Gridded
Clean up raw temperature data from Daymet gridded surface temperature query.
```{r mgmt gridded}
## Extract data ##
# Olympia #
oly.grid <- oly.grid$data
oly.grid$loc <- "Olympia"

# Port Townsend #
ptv.grid <- ptv.grid$data
ptv.grid$loc <- "Port Townsend"

# Toke Point #
tkp.grid <- tkp.grid$data
tkp.grid$loc <- "Toke Point"

# Neah Bay #
nhb.grid <- nhb.grid$data
nhb.grid$loc <- "Neah Bay"

# Clavert Island #
cvi.grid <- cvi.grid$data
cvi.grid$loc <- "Calvert Island"

## Combine ##
grid.dat <- rbind(oly.grid, ptv.grid, nhb.grid, tkp.grid, cvi.grid)

## Expand date data ##
grid.dat <- grid.dat %>%
  mutate(date = as.Date(paste(year, yday, sep = "-"), "%Y-%j")) %>% 
  mutate(month = format(date, "%m"), year = format(date, "%Y"), day = format(date, "%d"))

ggplot(grid.dat) +
  geom_line(aes(x = date, y = tmin..deg.c.)) +
  facet_wrap(~loc)

## Summarise summer data ##
summer.grid.summary <- grid.dat %>% 
  filter(month >= "06" & month <= "08") %>% 
  group_by(loc) %>% 
  summarise(avg.max = mean(tmax..deg.c.),
            sd.max = sd(tmax..deg.c.),
            n.max = n(),
            avg.min = mean(tmin..deg.c.),
            sd.min = sd(tmin..deg.c.),
            n.min = n()) %>% 
  mutate(se.max = sd.max / sqrt(n.max),
         lower.ci.max = avg.max - qt(1 - (0.05 / 2), n.max - 1) * se.max,
         upper.ci.max = avg.max + qt(1 - (0.05 / 2), n.max - 1) * se.max,
         se.min = sd.min / sqrt(n.min),
         lower.ci.min = avg.min - qt(1 - (0.05 / 2), n.min - 1) * se.min,
         upper.ci.min = avg.min + qt(1 - (0.05 / 2), n.min - 1) * se.min)

summer.grid.summary[1, 1] <- "Pruth Bay"
summer.grid.summary[3, 1] <- "Budd Inlet"
summer.grid.summary <- rbind(summer.grid.summary, summer.grid.summary[rep(4, 1), ])
summer.grid.summary[6, 1] <- "Vancouver"

summer.grid.summary$station.name.f <- factor(summer.grid.summary$loc, levels = c("Pruth Bay", "Vancouver", "Neah Bay", "Port Townsend", "Toke Point", "Budd Inlet"))

air.temp.redu.t <- merge(air.temp.redu, summer.grid.summary, by = "station.name.f", all.x = TRUE)
```

### Plots
#### Water Levels and Air Temp
```{r}
## Air Temperature ##
# Figure 3 #
ylab <- "Air temperature (°C)"
tiff("Figure3_AirTemp_Exposure_AllRegions.tif", height = 5600, width = 6300, units = "px", res = 800, compression = "lzw")
ggplot(air.temp.redu.t) +
  geom_line(aes(x = as.POSIXct(t), y = v), color = "#EB330E", size = 0.75) +
  geom_flame(aes(x = as.POSIXct(t), y = v, y2 = (avg.max + 2*sd.max))) +
  geom_vline(xintercept = as.POSIXct("2021-06-26 13:15:00"), linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct("2021-06-27 13:15:00"), linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct("2021-06-28 13:15:00"), linetype = "dashed") +
  annotate("rect", xmin = as.POSIXct("2021-06-24 00:00:00"), xmax = as.POSIXct("2021-06-24 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-24 21:11:00"), xmax = as.POSIXct("2021-06-25 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-25 21:11:00"), xmax = as.POSIXct("2021-06-26 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-26 21:11:00"), xmax = as.POSIXct("2021-06-27 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-27 21:11:00"), xmax = as.POSIXct("2021-06-28 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-28 21:11:00"), xmax = as.POSIXct("2021-06-29 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-29 21:11:00"), xmax = as.POSIXct("2021-06-30 05:12:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-30 21:11:00"), xmax = as.POSIXct("2021-07-01 00:00:00"), ymin = 0, ymax = 45, alpha = 0.1, fill = "black") +
  scale_y_continuous(limits = c(0, 45), breaks = seq(0, 45, by = 5)) +
  scale_x_datetime(breaks = seq(as.POSIXct("2021-06-24 00:00:00"), as.POSIXct("2021-07-01 00:00:00"), "1 day"), date_labels = "%d-%m", limits = c(as.POSIXct("2021-06-24 00:00:00"), as.POSIXct("2021-07-01 00:00:00"))) +
  xlab("Date") +
  ylab(ylab) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(.~station.name.f, nrow = 2, ncol = 3) +
  geom_hline(data = summer.grid.summary[, c(2, 3, 14)], aes(yintercept = avg.max), color = "#FFA703", size = 0.75) +
  geom_hline(data = summer.grid.summary[, c(5, 6, 14)], aes(yintercept = avg.min), color = "#03BCFF", size = 0.75) +
  geom_segment(data = water.level.exp, aes(y = 0, x = start, yend = 0, xend = end), color = "#EBD000", size = 3)
dev.off()

## Water levels ## 
# Figure S1 #
tiff("FigureS1_WaterLevels_AllRegions.tif", height = 5600, width = 3000, units = "px", res = 800, compression = "lzw")
ggplot(water.level.redu) +
  geom_line(aes(x = as.POSIXct(t), y = v), color = "#2B09D9", size = 0.75) +
  geom_vline(xintercept = as.POSIXct("2021-06-26 13:15:00"), linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct("2021-06-27 13:15:00"), linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct("2021-06-28 13:15:00"), linetype = "dashed") +
  annotate("rect", xmin = as.POSIXct("2021-06-24 00:00:00"), xmax = as.POSIXct("2021-06-24 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-24 21:11:00"), xmax = as.POSIXct("2021-06-25 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-25 21:11:00"), xmax = as.POSIXct("2021-06-26 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-26 21:11:00"), xmax = as.POSIXct("2021-06-27 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-27 21:11:00"), xmax = as.POSIXct("2021-06-28 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-28 21:11:00"), xmax = as.POSIXct("2021-06-29 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-29 21:11:00"), xmax = as.POSIXct("2021-06-30 05:12:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  annotate("rect", xmin = as.POSIXct("2021-06-30 21:11:00"), xmax = as.POSIXct("2021-07-01 00:00:00"), ymin = -1.5, ymax = 5, alpha = 0.1, fill = "black") +
  scale_y_continuous(limits = c(-1.5, 5), breaks = seq(-1, 5, by = 1)) +
  scale_x_datetime(breaks = seq(as.POSIXct("2021-06-24 00:00:00"), as.POSIXct("2021-07-01 00:00:00"), "1 day"), date_labels = "%d-%m-%Y", limits = c(as.POSIXct("2021-06-24 00:00:00"), as.POSIXct("2021-07-01 00:00:00"))) +
  xlab("Date") +
  ylab("Water level (m)") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~station.name.f, nrow = 6, ncol = 1)
dev.off()
```
